---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gdm

<!-- badges: start -->
<!-- badges: end -->

The `gdm` package provides function to fit, plot, summarize, and apply Generalized Dissimilarity Models.

### Citations

Fitzpatrick MC, Mokany K, Manion G, Ferrier S. (2021) gdm: Generalized Dissimilarity Modeling. R package version 1.5.0.

Mokany K, Fitzpatrick MC, Ware C, Woolley SNC, Ferrier S (in revision) A working guide to generalized dissimilarity modelling.

Ferrier S, Manion G, Elith J, Richardson, K (2007) Using generalized dissimilarity modelling to analyse and predict patterns of beta diversity in regional biodiversity assessment. Diversity & Distributions 13: 252-264.

## Installation

`gdm` is available on CRAN, and can be installed using:

``` r
install.packages("gdm")
```

## Examples

Forthcoming...

```{r example}
library(gdm)
## basic example code
```


